Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-05-17 analysis/paper/paper.Rmd
fix citation key

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationDentalPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationDentalPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).

>>>>

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationPlaque1991].

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
